Forest canopies are better generators of biodiversity in the tropics

### 2025-07-09 manuscript.Rmd
text and fig2 revisions

For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns from local to global scales [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a]. In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019]. The vertical nature of mountains provides more land surface per unit land area than flat regions and produces steep climate gradients over short distances that allow rapid rates of turnover. Combined, these factors generate an extraordinarily high biodiversity for a given area relative to flat and climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019]. Yet mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines that truncate the area of land suitable for diverse communities, and greater seasonal temperature variation, which reduces the climatic distinctiveness of elevational bands relative to tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013]. Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation [@macarthurBirdSpeciesDiversity1961]. Yet only recently has the three-dimensional nature of terrestrial landscapes been recognized as a critical dimension to our understanding of biogeographic patterns [@fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a] as our ability to measure vegetation structure and quantify its use by animals globally has increased with the collation of large ecological databases and the launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018; @mouraPhylogenyinformedCharacterisationGlobal2024a].

Tall, multi-layered forests generate biodiversity through similar mechanisms to mountains. By increasing habitat volume and producing vertical zones of microclimate and resource availability, tall forests allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. Thus, taller forests increase niche opportunities, which may allow additional species to join the assemblage and lead to greater richness within a given planimetric area compared to regions with shorter vertical vegetation structure [@xingEcologicalPatternsProcesses2023a; @macarthurBirdSpeciesDiversity1961; @macarthurLimitingSimilarityConvergence1967]. This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020]. Yet impacts of canopy height on richness are not consistent globally, with often weaker relationships in high latitude regions, which harbor fewer species than tropical forests with comparable vertical vegetation profiles [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020] - a pattern resembling that of mountains.

Vegetative complexity and climate may play key roles in mediating the relationships between canopy height and species richness by impacting the availability of vertical niche opportunities physiologically and constraining arboreal lifestyles [CITE]. Densely layered forest structures not only provide habitat across the forest profile, but produce vertical microclimate gradients via interactions with macroclimate conditions, solar radiation, and wind speed. From the humid and thermally stable understory to the dry and thermally variable canopy that experiences hot and cold temperature extremes [CITE], this vertical gradient may interact with latitudinal climate gradients to help or hinder the capacity for tall forests to generate high levels of biodiversity. For example, increased exposure to extreme cold in the canopies of high latitude forests may truncate the volume of habitat that can support arboreal species and thus limit richness despite tall canopies [@oliveiraVerticalStratificationInfluences2019]. Moreover, relatively low thermal variability in the canopies of tropical forests may increase the distinctness of vertical climate bands that support high turnover and thus enhance the capacity for tall forests to generate biodiversity in the tropics [CITE].

Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity [CITE], the mechanisms that allow tall forests to generate biodiversity remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and landuse change on biodiversity, and the conservation value of forests. To fill this gap, we combine range maps of nearly 24,000 terrestrial vertebrate species with the TetrapodTraits and EltonTraits databases to map species richness and assemblage verticality in wooded biomes across the globe. We first investigated the capacity for taller forests to support greater species richness in forests and less densely vegetated wooded biomes across latitudes and expected that tropical canopies would exhibit the fastest rates of increase in richness with canopy height. We then examined the mechanistic processes driving this pattern. If vertical stratification contributes to the capacity for tall forests to support diverse assemblages, we expect that species rich assemblages will exhibit high verticality (i.e., a high average vertical niche position) and that environmental factors limiting verticality will limit species richness. To test this hypothesis, we examined correlations between assemblage verticality and species richness. We then modeled assemblage verticality of the four vertebrate classes by macroclimate and vegetative structure and investigated the microclimatic processes that underlie these macroclimatic patterns. While three-dimensional climate data is not currently available globally given the large computational requirements of resolving micrometeorological processes within forest canopies, we modeled vertical microclimate profiles in six sites across the globe within tropical, temperate, and boreal forests by integrating forest structure data derived from NASA's Global Ecosystem Dynamics Investigation (GEDI) mission with high temporal resolution macroclimate data in a process-based microclimate model.

@@

To evaluate the capacity for tall canopies to generate biodiversity, we modeled vertebrate richness as a function of canopy height for all vertebrates and each vertebrate class globally and mapped the residuals. We then produced separate models for tropical, temperate, and boreal regions in forests and more sparsely vegetated habitats, including woodlands, savannas, and shrublands, to examine variation in the capacity for wooded habitats of the same height to generate biodiversity and variation in the rate of increasing richness with a given increase in canopy height across different regions.

@@

# The mediating effect of verticality
The vertical layering of species within forests may mediate the capacity for tall forests to support rich assemblages. By vertically stratifying across canopy layers, arboreal species can fill the larger volume of habitat offered by tall forests, increasing the richness of the assemblage [cite]. Within wooded ecosystems, we therefore expect species rich assemblages to exhibit high verticality, or a skew in the vertical distribution of species toward occurring in arboreal habitats [@oliveiraVerticalStratificationInfluences2019]. To quantify the verticality of an assemblage, we use the aforementioned trait databases, assigning verticality scores to each species based on their occurrence in fossorial (0), terrestrial (0.5), and arboreal (1) habitats [@scheffersVerticalArborealityHorizontal2017]. For species occurring in multiple habitats, we averaged scores across habitats. We then calculated verticality as the standard effect size of the average verticality score across species in an assemblage, using 100 simulations to determine if assemblages are more vertical than expected by random chance given their richness and the evolutionary history of the biogeographic realm [@oliveiraVerticalStratificationInfluences2019] (Supplementary methods).

Verticality is strongly correlated with richness patterns, peaking in the tropics and declining toward higher latitudes.
- At its highest, x % of species exhibited arboreal strategies, corresponding to some of the richest assemblages in tropical regions.
- In these regions, expansion up the canopy combined with fine-scale niche partitioning, and thus species packing, within the canopy produce species rich assemblages.

Though temperate and boreal regions have tall trees, albeit not quite reaching the heights of the tallest tropical canopies, they contain assemblages with much lower verticality, consisting of greater proportions of fossorial and terrestrial species. This pattern suggests that the capacity for taller canopies to support higher richness declines when canopies cannot support highly vertical species assemblages.

Environmental factors that constrain the evolution of diverse arboreal assemblages are therefore key to unraveling the vertical dimension to Humbolt's enigma.

constraining verticality or the fine-scale partitioning of niche space will limit the capacity for tall forests to generate biodiversity.

# Canopy Climates
To examine the mechanisms contributing to variation in the capacity for tall forests to support highly arboreal, and thus species rich, assemblages, we use spatial linear mixed models to model assemblage verticality for each vertebrate class as a function of canopy height and its interaction with climatic factors that may constrain or support arboreality, including minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the warmest month (Supplementary methods). We additionally include vegetation density in the models, as this will impact the availability of habitat volume, and did not include foliage height diversity due to its strong correlation with canopy height, as has previously been observed [@decontoCharacterizingStructuralComplexity2024; @atkinsPowerLawScaling2022]. The climate variables included in these models represent macroclimate conditions in open habitats rather than the climate conditions experienced in forest canopies. We therefore additionally modeled vertical microclimate profiles in six lowland forests from tropical to boreal latitudes by integrating information on the vertical distribution of vegetation derived from NASA's GEDI spaceborne lidar mission [] with ERA5 macroclimate data at a high temporal resolution [] in a process-based microclimate model that resolves within-canopy processes generating vertical climate gradients []. These models provide an estimate of climate conditions proximal to those experienced by forest-dwelling organisms and thus better represent the biogeographic and ecophysiological processes that allow tall forests to support vertical and rich assemblages in the tropics more so than at higher latitudes [@klingesProximalMicroclimateMoving2024]. Modeling microclimates in conjuction with modeling verticality based on macroclimate conditions thus allows us to infer the microclimate processes that underlie macroclimate patterns in the absence of global 3D climate data.

Our models indicate that taller canopies support highly vertical vertebrate assemblages, while dry, hot, and cold climates limit verticality and mediate the capacity for tall forests to support vertical assemblages, albeit to varying extents across vertebrate classes (Fig. \@ref(fig:fig3). These patterns arise because the interaction between macro- and micro-climate gradients reduce opportunities for niche specialization across vertical strata via impacts on vegetation structure and place physiological constraints on arboreal lifestyles.

These patterns arise because macroclimate indirectly constrain verticality via its effect on vegetation structure and niche opportunities, while microclimate conditions produced by the interaction between macroclimate and vegetation structure place physiological constraints on arboreal lifestyles.  the interaction between macro- and micro-climate gradients reduce opportunities for niche specialization across vertical strata via impacts on vegetation structure and place physiological constraints on arboreal lifestyles.

This occurs because climate indirectly constrains verticality via its effects on vegetation structure and niche opportunities for specialization across vertical strata and the interaction between macro- and micro-climate gradients directly constrain verticality....

Climate mediates that capacity for forests to support vertical assemblages indirectly, through its effect on vegetation structure, and directly, via impacts on animal physiology.

Interactions between macroclimate, vegetation structure, and microclimate impact verticality by affecting niche opportunities and imposing physiological constraints.

Depends on the distribution of vegetation as well as the spatial and temporal variation of climate in the canopy.

## Climate produces structural constraints to verticality
Climate impacts the structural complexity of vegetation, which plays a key role in the the availability of arboreal niche opportunities and is typically associated with high animal richness [@xingEcologicalPatternsProcesses2023a; @daviesAdvancesAnimalEcology2014]. High annual precipitation and low precipitation seasonality enable development of tall and structurally complex forests with densely layered vegetation and generally high tree diversity [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024; @hakkenbergClimateMediatesRelationship2021]. Wet climates additionally allow development of diverse epiphyte communities, which provide additional microhabitats in the canopy [CITE]. Hotspots of structural complexity that coincide with some of the tallest forests, including the Choco-Darien, the Amazon, Borneo, and New Guinea [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024], thus provide opportunities for fine-scale niche partitioning, yielding extraordinarily rich communities. In contrast, lower structural complexity in parts the Afrotropics [@ehbrechtGlobalPatternsClimatic2021a] may reduce niche opportunities for arboreal species, yielding less vertical assemblages and lower richness than expected based on the height of the canopy (Fig. \@ref(fig:fig1)).

At higher latitudes, lower precipitation and light hinder the development of densely layered forests by restricting the functional and structural configurations of trees [@ehbrechtGlobalPatternsClimatic2021a; @hakkenbergClimateMediatesRelationship2021; @atkinsPowerLawScaling2022]. In evergreen needleleaf forests that dominate boreal regions, these environmental constraints dampen the rate of increase in structural complexity with canopy height [@atkinsPowerLawScaling2022]. Tall canopies therefore provide fewer niche opportunities in boreal regions, which may constrain verticality and limit species richness relative to canopies of equivalent height in lower latitude forests. Yet, hotspots of structural complexity in temperate forests, including the North Pacific Alaskan coastal forest, Valdivian forest, and Tasmanian forest, that nearly match the complexity of tropical forests [@ehbrechtGlobalPatternsClimatic2021a] also contain assemblages with lower verticality than the tropics and fewer species than expected based upon their canopy height (Figs. \@ref(fig:fig1), \@ref(fig:fig2)). MacArthur & MacArthur [-@macarthurBirdSpeciesDiversity1961] observed this pattern over 50 years ago, noting that "tropical bird species diversity is much greater than the temperate one for habitats of comparable profiles." The effects of the interaction between macroclimate and vegetation profiles on vertical microclimate profiles provide an explanation [MIRRORING MECHANISMS THAT ALLOW MOUNTAINS TO BE BETTER GENERATORS OF BIODIVERSITY IN THE TROPICS]

## Physiological constraints to arboreality truncate suitable habitat 
## Canopy climates truncate suitable habitat in cold and dry forests

Spatial and temporal climate variability from macro to micro scales strongly impact macroecological patterns (Klinges and Scheffers 2021; Rahbek)  [CITE].
Vertical features in the landscape, including mountains and trees bring distinct climates within close proximity of each other [@rahbekHumboldtsEnigmaWhat2019]. By absorbing and reflecting solar radiation and reducing vertical air mixing, dense forests produce steep vertical microclimate gradients [THAT CAN CHANGE AT RATES OF UP TO XXX PER METER] and therefore capture a remarkable volume of climatic variation given their geographic footprint. For example, canopies experience drier conditions and hotter maximum temperatures than understories across latitudes, while temperate and boreal canopies additionally experience colder minimum temperatures [FIGURE AND SUPP FIGURES WITH ADDED DETAIL]. When macroclimate conditions producing canopy microclimates favorable for arboreal life, species may partition climate space across the entire forest volume. For example, high dry season precipitation and thus low precipitation seasonality in tropical moist forests support arboreal life by limiting seasonal water stress in the canopy and allow arboreal animals to maintain body temperatures below critical thermal maxima via evapotranspiration while avoiding desiccation [@mckechnieAvianThermoregulationHeat2017; @sundayThermalsafetyMarginsNecessity2014] (Fig. \@ref(fig:fig3)). Low physiological stress allows high rates of turnover across the vertical gradient that yields a dense packing of species in the canopy that increases assemblage verticality and richness [CITE].

In contrast, unfavourable canopy climates may reduce the number of viable physiological strategies thereby limiting the extent to which taller canopies can support increasing richness. For example, low dry season precipitation, which inhibits verticality (Fig. \@ref(fig:fig3)), exaggerates water stress in canopies and may prevent fully arboreal species from occurring. In tropical dry forests, animals seasonally adjust habitat use, shifting from using primarily arboreal habitats during the wet season to more humid terrestrial habitats during the dry season to avoid desiccation (Basham and Scheffers 2020; Rajaonarivelo et al. 2021).  In temperate and boreal biomes, low precipitation combined with dry canopy microclimates may similarly constrain ectotherm arboreality, where basking to increase body temperatures can quickly lead to desiccation without adequate moisture availability or physiological defenses (Sunday et al. 2014; Tracy, Christian, and Tracy 2010). Harsh macroclimate conditions therefore seasonally truncate the volume of habitat that can support arboreal species, which collapses vertically stratified assemblages and reduces opportunities for fine-scale partitioning of the entire vertical climate gradient. [And flattens rates of turnover as height increases??] This prevents taller canopies from supporting the addition of arboreal species to assemblages.

PHYSIOLOGICAL TOLERANCE HYPOTHESIS - REDUCE NUMBER OF FUNCTIONAL TYPES THAT CAN INHABIT A CLIMATE SPACE, LIMITING THE EXTENT TO WHICH AN INCREASE IN CANOPY HEIGHT INCREASES SPECIES RICHNESS
ENVIRONMENTAL FILTERING IN ARBOREAL COMMUNITIES LIMITS ARBOREAL DIVERSITY IN HARSHER MACROCLIMATE CONDITIONS

Harsh macroclimate conditions place environmental filters on viable physiologies for arboreal species, effectively truncating the viable functional space and therefore the amount of habitat volume that can be occupied.
[https://nsojournals.onlinelibrary.wiley.com/doi/full/10.1111/ecog.07122]

>>>>

For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns from local to global scales [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a]. In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019]. [Yet until recently, hypotheses failed to explain the extraordinary richness of mountains particularly in the tropics.] The vertical nature of mountains provides more land surface per unit area than flat regions and produces steep climate gradients over short distances that allow rapid rates of species turnover. Combined, these factors allow mountains to generate an extraordinarily high biodiversity for a given area relative to flat and climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019]. Yet mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines that truncate the area of land suitable for diverse communities, and greater seasonal temperature variation, which reduces the climatic distinctness of elevational bands relative to tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013].

NEED TO MENTION HUMBOLT'S ENIGMA SOMEWHERE IN FIRST PARAGRAPH AND THAT THE MAIN EXPLANATIONS ARE TOPOGRAPHY AND CLIMATE

Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation, which is increasingly being recognized as critical to our understanding of biogegoraphic patterns [@macarthurBirdSpeciesDiversity1961]. Tall, multi-layered forests generate biodiversity through similar mechanisms to mountains. By increasing habitat volume and producing vertical zones of microclimate and resource availability, tall forests allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. Thus, taller forests increase niche opportunities, which may allow additional species to join the assemblage and lead to greater richness within a given planimetric area compared to regions with shorter vertical vegetation structure [@xingEcologicalPatternsProcesses2023a; @macarthurBirdSpeciesDiversity1961; @macarthurLimitingSimilarityConvergence1967]. This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020]. Yet positive impacts of canopy height on richness are not consistent globally, with often weaker relationships in high latitude regions where forests of comparable vertical profiles to those in the tropics harbor fewer species [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020]. - a vertical dimension to Humbolt's enigma

Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity [@macarthurBirdSpeciesDiversity1961; @scheffersIncreasingArborealityAltitude2013; @xingEcologicalPatternsProcesses2023a], the mechanisms that help or hinder the capacity for tall forests to generate biodiversity remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and landuse change on biodiversity, and the conservation value of forests. Following upon explanations to Humbolt's enigma [@rahbekHumboldtsEnigmaWhat2019], we fill this gap by examining how vegetative complexity and climate may play key roles in mediating the relationship between canopy height and species richness by impacting the capacity for tall canopies to support vertically stratified and species rich arboreal assemblages. For example, the denser layering of vegetation in tropical forests may provide more niche opportunities, allowing tall canopies to support more biodiversity than canopies of equivalent height in temperate regions [CITE or CONCEPT FIG]. Moreover, densely layered vegetation produces vertical microclimate gradients via interactions with macroclimate conditions, solar radiation, and wind speed that are characterized by humid and thermally stable understories to dry and thermally variable canopies that experiences both hot and cold temperature extremes. Physiological constraints imposed on arboreal life strategies by these more extreme conditions, combined with the distinctness of vertical climate bands may impact species turnover from the ground to the canopy and thus the capacity for taller canopies to support increasing richness [@bashamVerticalStratificationCollapses2020] [CITE]. Yet only recently have we been able to address these hypotheses, as our ability to measure vegetation structure and quantify its use by animals globally [AND MODEL MICROCLIMATES GLOBALLY] increased with the collation of large ecological databases and the launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018; @mouraPhylogenyinformedCharacterisationGlobal2024a].

In this study we combine range maps of nearly 24,000 terrestrial vertebrate species and comprehensive databases of vertical habitat use, global macroclimate and forest structure data, and process-based microclimate models to examine how forest structure and climate may mediate the capacity for tall forests to support species rich assemblages via impacts on the vertical structure of vertebrate communities. We first investigated the capacity for taller forests to support increasing species richness in forests and less densely vegetated wooded biomes across latitudes. If the vertical structure of species assemblages mediates the relationship between canopy height and species richness, then we expect positive correlations between the verticality of assemblages (i.e., their average vertical niche position) and species richness. [AND THAT CANOPY HEIGHT WILL EXPLAIN LITTLE VARIATION IN RICHNESS ONCE VERTICALITY HAS BEEN ACCOUNTED FOR]

and expected that tropical canopies would exhibit the fastest rates of increase in richness with canopy height. We then examined the mechanistic processes driving this pattern.  EXPLANATION OF TETRAPODTRAITS AND ELTONTRAITS

, including TetrapodTraits and EltonTraits, to map species richness and the vertical structure of species assemblages in wooded biomes across the globe.

To fill this gap, we combine range maps of nearly 24,000 terrestrial vertebrate species with the TetrapodTraits and EltonTraits databases to map species richness and the vertical structure of species assemblages in wooded biomes across the globe. We first investigated the capacity for taller forests to support greater species richness in forests and less densely vegetated wooded biomes across latitudes and expected that tropical canopies would exhibit the fastest rates of increase in richness with canopy height. We then examined the mechanistic processes driving this pattern. If vertical stratification contributes to the capacity for tall forests to support diverse assemblages, we expect that species rich assemblages will exhibit high verticality (i.e., a high average vertical niche position) and that environmental factors limiting verticality will limit species richness. To test this hypothesis, we examined correlations between assemblage verticality and species richness. We then modeled assemblage verticality of the four vertebrate classes by macroclimate and vegetative structure and investigated the microclimatic processes that underlie these macroclimatic patterns. While three-dimensional climate data is not currently available globally given the large computational requirements of resolving micrometeorological processes within forest canopies, we modeled vertical microclimate profiles in six sites across the globe within tropical, temperate, and boreal forests by integrating forest structure data derived from NASA's Global Ecosystem Dynamics Investigation (GEDI) mission with high temporal resolution macroclimate data in a process-based microclimate model.

@@

To evaluate the capacity for tall canopies to generate biodiversity, we modeled vertebrate richness as a function of canopy height for all vertebrates and each vertebrate class globally. We then mapped the residuals of these models to determine where canopy height over- or under-predicted species richness. We then produced separate models for tropical, temperate, and boreal regions in forests and more sparsely vegetated habitats, including woodlands, savannas, and shrublands, to examine variation in the capacity for wooded habitats of the same height to generate biodiversity and variation in the rate of increasing richness with a given increase in canopy height across different regions.

@@

# The mediating effect of verticality
The vertical layering of species within forests may mediate the capacity for tall forests to support rich assemblages. As forests increase in height, additional arboreal species may be added to the assemblage in novel canopy space [@macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @larueDiversityVolumeRelationships2023]. An increase in the number of arboreal species without an equivalent increase in terrestrial or fossorial species increases the verticality of the assemblage, which represents the skew in the vertical distribution of species toward the canopy [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017]. Within wooded ecosystems, we therefore expect regions exhibiting high verticality to support species rich assemblages and steep increases in richness with canopy height [@oliveiraVerticalStratificationInfluences2019]. To quantify the verticality of an assemblage, we use the aforementioned trait databases, assigning verticality scores to each species based on their occurrence in fossorial (0), terrestrial (0.5), and arboreal (1) habitats [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017]. For species occurring in multiple habitats, we averaged scores across habitats. We then calculated verticality as the standard effect size of the average verticality score across species in an assemblage, using 100 simulations to determine if assemblages are more vertical than expected by random chance given their richness and the evolutionary history of the biogeographic realm [@oliveiraVerticalStratificationInfluences2019] (Supplementary methods).

Verticality peaked in the tropics and declined toward the poles, exhibiting strong correlations with verticality across taxa (Fig. \@ref(fig:fig2)). Moist tropical forests contained assemblages comprised of x% to x% species exhibiting arboreal strategies (SUPP FIG), corresponding to some of the most species rich assemblages. Though temperate and boreal regions contain tall forests, albeit not quite reaching the heights of the tallest tropical canopies, they contain assemblages with much lower verticality, and higher proportions of species exhibiting terrestrial and fossorial strategies. This pattern suggests that the capacity for taller canopies to support higher richness declines when canopies cannot support highly vertical species assemblages. Understanding environmental factors that constrain the evolution of diverse arboreal assemblages is therefore key to unraveling the vertical dimension to Humbolt's enigma.

# Vegetation structure and canopy climates impact the capacity for tall canopies to support vertical communities
To examine the environmental factors influencing assemblage verticality, we used spatial linear mixed models to model assemblage verticality for each vertebrate class as a function of canopy height and its interaction with macroclimatic factors that may constrain or support arboreal life strategies, including minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the warmest month, as well as with macroclimatic factors that may influence the distinctness of vertical climate banding, including diurnal temperature range (Supplementary methods). We additionally included vegetation density in the models, which impacts the availability of habitat volume. While foliage height diversity and temperature seasonality also impact niche opportunities and the distinctness of climate banding, these variables where strongly correlated with canopy height and minimum temperatures, respectively, and were therefore excluded from the models. The macroclimate variables included in these models represent climate conditions experienced in open habitats rather than the microclimate conditions across forest profiles [@defrenneWeatherStationsLack2016; @bramerAdvancesMonitoringModelling2018]. While using macroclimate data was necessary due to the lack of global vertically resolved microclimate data, its use inhibits our ability to understand how microclimates experienced by forest-dwelling organisms impact community structure [@klingesProximalMicroclimateMoving2024; @zellwegerForestMicroclimateDynamics2020]. We therefore additionally modeled vertical microclimate profiles in six forests from tropical to boreal latitudes by integrating information on the vertical distribution of vegetation derived from NASA's GEDI spaceborne lidar mission [@dubayahGlobalEcosystemDynamics2020] with ERA5 macroclimate data at a high temporal resolution [@hersbachERA5HourlyData2023] in a process-based microclimate model that resolves within-canopy processes to predict vertical climate gradients [@macleanMicroclimfFastCanopy2024]. To reduce variability in vertical profiles not attributed to latitudinal differences, we restricted microclimate models to points below 500 m a.s.l and that represented the top 75th to 98th quantiles of canopy height within a 1° x 1° grid cell. By estimating the climate conditions experienced by forest-dwelling organisms, these models provide insight into the microclimatic processes that underlie macroclimate patterns and allow canopies to support vertically stratified and species rich assemblages.

Our models of verticality indicate that taller canopies support highly vertical vertebrate assemblages, while dry, and thermally variable macroclimate conditions, including extreme hot and cold temperatures limit verticality and mediate the capacity for tall forests to support vertical assemblages, albeit to varying degrees across vertebrate classes (Fig. \@ref(fig:fig3)). These patterns arise due to impacts of spatial climate variation on vegetation structure and physiological constraints to arboreal lifestyles as well as impacts of temporal climate variation on the distinctness of vertical climate bands.

## Climate produces structural constraints to verticality
Climate impacts the structural complexity of vegetation, which plays a key role in the the availability of arboreal niche opportunities and is typically associated with high animal richness [@xingEcologicalPatternsProcesses2023a; @daviesAdvancesAnimalEcology2014]. High dry season precipitation and thus low precipitation seasonality may support highly vertical assemblages by enabling development of tall and structurally complex forests with densely layered vegetation and generally high tree diversity [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024; @hakkenbergClimateMediatesRelationship2021]. Wet climates also support diverse epiphyte communities, which provide additional microhabitats in the canopy that contribute to high arboreal richness [@taylorVascularEpiphytesContribute2022; @thomsenSecondaryFoundationSpecies2018; @ellwoodDoublingEstimateInvertebrate2004]. Hotspots of structural complexity that coincide with some of the tallest forests, including the Choco-Darien, the Amazon, Borneo, and New Guinea [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024], thus provide opportunities for fine-scale niche partitioning, yielding species rich assemblages [@camargoForestVerticalComplexity2018; @gouveiaForestStructureDrives2014; @oliveiraVerticalStratificationInfluences2019]. In contrast, lower structural complexity in parts the Afrotropics [@ehbrechtGlobalPatternsClimatic2021a] may reduce niche opportunities for arboreal species, yielding less vertical assemblages and lower richness than expected based on the height of the canopy (Fig. \@ref(fig:fig1)).

At higher latitudes, lower precipitation and light hinder the development of densely layered forests by restricting the functional and structural configurations of trees [@ehbrechtGlobalPatternsClimatic2021a; @hakkenbergClimateMediatesRelationship2021; @atkinsPowerLawScaling2022]. In evergreen needleleaf forests that dominate boreal regions, these environmental constraints dampen the rate of increase in structural complexity with canopy height [@atkinsPowerLawScaling2022]. Tall canopies therefore provide fewer niche opportunities in boreal regions, which may constrain assemblage verticality and limit species richness relative to canopies of equivalent height at lower latitudes. Yet, hotspots of structural complexity in temperate forests, including North Pacific Alaskan, Valdivian, and Tasmanian forests, that nearly match the complexity of tropical forests [@ehbrechtGlobalPatternsClimatic2021a] also contain assemblages with lower verticality than the tropics and fewer species than expected based upon their canopy height (Figs. \@ref(fig:fig1), \@ref(fig:fig2)). MacArthur & MacArthur [-@macarthurBirdSpeciesDiversity1961] observed this pattern over 50 years ago, noting that "tropical bird species diversity is much greater than the temperate one for habitats of comparable profiles." The interaction between macroclimate and vertical microclimate profiles provides an explanation that mirrors the climatic mechanisms allowing mountains to be better generators of biodiversity in the tropics [@rahbekHumboldtsEnigmaWhat2019].

## The influence of canopy climates

Vertical features in the landscape, including mountains and trees bring distinct climates within close proximity of each other [@rahbekHumboldtsEnigmaWhat2019; @leahyVerticalNicheElevation2021; @vinodThermalSensitivityForest2023]. By absorbing and reflecting solar radiation and reducing vertical air mixing, dense forests produce vertical microclimate gradients that capture a remarkable volume of climatic variation given their geographic footprint [@defrenneForestMicroclimatesClimate2021]. These steep climate gradients can support high rates of turnover from the ground to the canopy that yield vertically stratified and species rich assemblages [@bashamVerticalStratificationCollapses2020; @xingSpeciesTurnoverAnt2022; @oliveiraVerticalStratificationInfluences2019; @camargoForestVerticalComplexity2018]. However, more extreme microclimates in the canopy than the understory may place physiological constraints on arboreal lifestyles depending on background macroclimate conditions, effectively truncating the available volume of forest and limiting the capacity for tall forests to support high richness, just as lower elevation snow lines limit the capacity for tall mountains to generate biodiversity at higher latitudes [@oliveiraVerticalStratificationInfluences2019].

For example, our models indicate that low dry season precipitation constrains verticality across vertebrate classes (Fig. \@ref(fig:fig3)). These macroclimate patterns were accompanied by steep gradients in minimum relative humidity, which declined x% from the ground to the top of the canopy (Fig. \@ref(fig:fig4)). Low precipitation may exaggerate the drier canopy conditions and risk desiccation in hot climates, where arboreal species depend on evapotranspiration to maintain body temperatures below critical thermal maxima, or in cold climates, where basking to increase body temperature can quickly lead to desiccation without adequate moisture availability or physiological defenses. By imposing environmental filters on the functional morphologies that can inhabit canopy environments, dry climates therefore reduce assemblage verticality and limit the number of species that an increase in canopy height can support, in line with the physiological tolerance hypothesis [https://onlinelibrary.wiley.com/doi/full/10.1111/j.1461-0248.2004.00671.x]. This pattern is evident for frog communities in the seasonally dry forests of Panama, where collapse of vertical stratification toward the ground during the dry season due to water limitation in the canopy is accompanied by richness declines and homogenization of arboreal diversity [@bashamVerticalStratificationCollapses2020]. Simply put, extreme canopy microclimates that physiologicall constrain arboreal life truncate the volume of available habitat, which prevents tall trees from supporting species rich assemblages. In contrast, high dry season precipitation in tropical moist forests supports highly vertical assemblages by reducing physiological constraints on arboreal life strategies in the upper canopy. Species are therefore able to finely partition the entire volume of forested habitat, leading to a dense packing of species in the canopy and allowing increases in canopy height to yield rapid increases in species richness (Fig. \@ref(fig:fig1)).

Hot and cold temperatures had less consistent effects across taxa, but may still play critical roles in physiological constraints, particularly in regions exhibiting steeper vertical gradients characterized by hotter maximum and colder minimum temperatures. For example, high maximum temperatures reduced verticality and/or [REDUCED THE POSITIVE IMPACT OF CANOPY HEIGHT ON VERTICALITY FOR REPTILES, MAMMALS, AND BIRDS] across vertebrate classes, while low minimum temperatures reduced positive effects of canopy height for birds and reptiles (Fig. \@ref(fig:fig3)). High maximum temperatures may cause ectotherm body temperatures to exceed critical thermal maxima (McKechnie et al. 2017; Sunday et al. 2014), particularly in drier regions where maximum temperatures peaked. Microclimate models additionally indicated that vertical maximum temperature gradients were steepest in the temperate forests, though canopy temperatures exceeding 40C may limit the capacity for species to stratify across the entire vertical climate gradient (Fig. \@ref(fig:fig4)). Similarly, low minimum temperatures can cause body temperatures to exceed critical thermal minima for ectotherms or impose energetic constraints to endotherms, particularly in the windier conditions that characterize the canopy  (Sunday et al. 2014) (Byman, Hay, and Bakken 1988). Shallow vertical minimum temperature gradients (Fig. \@ref(fig:fig4)), particularly during leaf off conditions [CITE], cause environmental filters to act across the entire canopy, posing severe limitations on the capacity for taller canopies to increase species richness.

[https://nsojournals.onlinelibrary.wiley.com/doi/full/10.1111/ecog.07122]
